MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-03-08 Article.tex
author change
major revision

To showcase the utility of the new implementations, we performed benchmarking on both existing models and large-scale synthetic models. As the main results, MaBoSS.GPU may provided over 200\texttimes\ speedup over the current version of MaBoSS on a wide range of models using contemporary GPU accelerators, and MaBoSS.MPI is capable of almost linear performance scaling with added HPC resources, allowing similar speed-ups by utilizing the current HPC infrastructures.

@@

\subsection{Computational complexity of MaBoSS algorithm in parallel}

@@

% TODO this paragraph needs friggin references to the algorithm figure -- took me a while to see where to get the "all other parts"
We estimate the time required to simulate $c$ trajectories as follows:
For simplification, we assume that a typical Boolean logic formula in a model of $n$ nodes can be evaluated in $\mathcal{O}(n)$ (this is a very optimistic but empirically valid estimate). With that, the computation of all possible transition rates can be finished in $\mathcal{O}(n^2)$. The selection of the flipping bit can be finished in $\mathcal{O}(n)$, and all other parts of the iteration can finish in $\mathcal{O}(1)$. In total, the time complexity of one iteration is $\mathcal{O}(n^2)$. If we simulate $c$ trajectories with an upper bound of trajectory length $u$, the simulation time is in $\mathcal{O}(c \cdot u \cdot n^2)$.

In an idealized PRAM model with infinite parallelism, we can optimize the algorithm in the following ways:
\begin{itemize}
    \item Given $c$ processors, all trajectory simulations can be performed in parallel, reducing the time complexity to $\mathcal{O}(m \cdot n^2)$.
    %TODO might be useful to explicitly note that this doesn't include result aggregation (and link to the section)
    \item With $n$ processors, the computation of transition rates in the simulation can be done $\mathcal{O}(n)$ time, and the selection of the flipping bit can be done in $\mathcal{O}(\log{n})$ time using a parallel prefix sum, giving $\mathcal{O}(n)$ time for a single iteration.
\end{itemize}
Thus, using a perfect parallel machine with $c \cdot n$ processors, the computation time can be reduced to $\mathcal{O}(u \cdot n)$. Notably, the $\mathcal{O}(u)$ simulation steps that must be performed serially remain a major factor in the whole computation time.

@@

In the CPU version of MaBoSS, the simulation part is the most computationally demanding part, with up to 80\% of MaBoSS runtime spent by just evaluating the Boolean formulas (the exact number depends on the model). The original formula evaluation algorithm in MaBoSS used a recursive traversal of the expression tree, which (apart from other issue) causes memory usage patterns unsuitable for GPUs: the memory required per each core is not achievable in current GPUs, and there are typically too many cache misses~\cite{karlsson2000prefetching}

@@

For optimizing the statistics aggregation, MaBoSS.GPU heavily relies on the fact that the typical number of non-internal nodes in a real-world MaBoSS model rarely exceeds 10 nodes, regardless of the size of the model. This relatively low number of states generated by non-internal nodes allows us to materialize the whole statistics structure (called ``histogram'') as a fixed-size array (rarely exceeding $2^10$ elements).

This approach allows us to avoid storing the states as the keys, and gives a simple approach that can map the state to the histogram index using simple bit masking and shifting instructions. Further, we use sevelra well-known GPU histogram update optimizations to improve the performance, including as shared memory privatization and atomic operations.

@@

% TODO: ok well, here you have a grande totale insane of 3 models, and decide to omit the size of the middle one by saying "yeah between 10 and 133" :D
For the benchmarks, we used $3$ real-world models of $10$, $XXX$ and $133$ nodes (\texttt{cellcycle}~\cite{faure2006cellcycle}, \texttt{sizek}~\cite{sizek2019boolean} and \texttt{Montagud}~\cite{montagud2022prostate}). In order to test the scalability of the GPU and MPI implementation, we also created several synthetic models with up to $1000$ nodes. Synthetic models were designed in a way such that the length of each trajectory is predictable, and the models have no stable states. Also, the number of non-internal nodes was kept low (10 nodes) to enable the usage of the histogram optimization.
%TODO here it would be nice to calm down stupid questions like "ok how realistic are the models actually" by pointing out that they are in the repository or something.

The GPU implementation benchmarks were run on a datacenter-grade NVIDIA Tesla A100 GPU and a consumer-grade NVIDIA RTX 3070 Laptop GPU. The CPU implementation benchmarks were run on a 32-core Intel Xeon Gold 6130 CPU with multithreading. The CPU implementation was compiled with GCC 13.2.0, and the GPU implementation was compiled with CUDA 12.2. Each measurement was repeated $10$ times, and the average runtime was used as the final results after removing the outliers.
%TODO: stats trigger: how did you remove the outliers? Manual QA? What caused the outliers?

The MPI implementation benchmarks were run on the MareNostrum 4 supercomputer.
% TODO link to MN spec (I think they have a paper)

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{Figures/real.pdf}
    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on real-world models. Each model is simulated with 1 million trajectories.}
    % TODO it's slightly inconvenient that the plot legend only describes hardware; in turn the whole thing looks like you're benchmarking the hardware instead of the actual sofware implementations. Suggest using: MaBoSS (CPU), MaBoSS.GPU (RTX3070), MaBoSS.GPU (A100).
    \label{fig:real}
\end{figure}

% TODO above you're putting the Montagud/Sizek/whatever names into \texttt. I'd recommend using \textsc and doing it veeery consistently everywhere. Just for the general purpose of sparking the benchmark-nerd imaginations.
In Figure~\ref{fig:real}, we compare the wall time of the CPU and GPU implementations on real-world datasets. The GPU implementation is faster than the CPU implementation on all models, and the speedup shows to be more significant on the models with more nodes and longer trajectories. On the Montagud model with 133 nodes, but a relatively short average trajectory, we achieve $145\times$ speedup. On a slightly smaller Sizek model with a longer average trajectory, the speedup is up to $326\times$.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{Figures/nodes.pdf}
    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on synthetic models. Each model is simulated with 1 million trajectories. The two panels differ by inclusion of the runtime GPU code compilation, showing its impact on total run time.}
    \label{fig:synth}
\end{figure}

Figure~\ref{fig:synth} shows much finer performance progression on synthetic models. We observed that the CPU variant starts to progress steeper at around the $100$ nodes boundary. We assume that the implementation hits the cache size limit, and the overhead of fetching the required data from the memory becomes dominant. The same can be observed in the GPU variant later at around $200$ nodes. Expectably, the cache-spilling performance penalty is much more significant on GPUs. Overall, the results suggest that the optimization of dividing transition rate computations among multiple threads, as mentioned in the previous section XXX TODO REF XXX, may provide a better speedup for bigger models, as it alleviates the register and cache pressure.
% TODO danger there's a non-commented TODO above. Add an actual \ref if possible.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{Figures/nodes-compilation-big-NVIDIA Tesla A100 GPU.pdf}
    \caption{The time spent in the runtime compilation of the Boolean formulas simulating models with varying numbers of nodes, trajectories, and formula lengths.}
    % TODO: what happens if we swap the sample count (color) with formula length (facet) ?
    % Also, it might be much better to have the percentages as logarithmic scale. (most of the space is now killed by the single 200node 10formula point)
    % TODO: "sample count" in the legend is not defined anywhere around, maybe say "total trajectories" ?
    \label{fig:comp}
\end{figure}

Additionally, Figure~\ref{fig:synt} shows the total runtime of the GPU implementation including the runtime compilation step. Comparing the panels, we observe that that the relative runtime compilation overhead quickly disappears with increasing model size. Figure~\ref{fig:comp} shows the results of a more detailed benchmarks for this scenario, as run on the NVIDIA Tesla A100 GPU. We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes). Notably, as soon as the simulation becomes more complex (e.g., by increasing the number of nodes or simulated trajectories), the compilation time becomes relatively negligible even for models with unrealistically long formulae. This suggests that the runtime compilation is a viable optimization methodology also for much larger models.

@@

%TODO achtung, the figure DPI is somehow messed up and the letters are too tiny.
%TODO: might be better to merge the figures into one (2 facets)
%TODO: it's nice that there's a trend on the diagonal but that kinda looks like we're hiding fluctuations or small off-angle, can we scale the thing into a visibly horizontal straight line?
%  - multiply the walltime by the number of cores to produce the realistic totalwastedtime
%  - divide the speedup by the number of cores invested (to show speedup per investment)

\begin{figure}%[tbhp]
\centering
\includegraphics[width=3.25in]{Figures/sizek_mpi.pdf}
\caption{Scalability results of MPI implementation on Sizek model with 20 cores per MPI node.}
\label{fig:sizek_results}
\end{figure}

Figure~\ref{fig:sizek_results} shows the efficiency of the MaBoSS.MPI implementation on the Sizek model. We ran multiple suites, ranging from a single MPI node up to 192 nodes, each running 20 cores. We can observe a close-to-linear speedup of up to 64 MPI nodes (1280 cores), and a plateau for larger suites (Figure~\ref{fig:sizek_results}, green). This can be explained by hitting an expectable bottleneck in parallelization overhead and MPI communication cost when the problem is divided into too many small parts.

\begin{figure}%[tbhp]
\centering
\includegraphics[width=3.25in]{Figures/synth_mpi_speedup.pdf}
\caption{Speedup scaling of MPI implementation on a synthetic model with 1000 nodes and 32 cores per MPI node.}
\label{fig:synthetic_results}
\end{figure}

@@

% \begin{figure}%[tbhp]
% \centering
% \includegraphics[width=.8\linewidth]{Figures/large_model.pdf}
% \caption{Scalability results for synthetic model with 1000 nodes}
% \label{fig:synthetic_results}
% \end{figure}

%\begin{figure}%[tbhp]
%\centering
%\includegraphics[width=.8\linewidth]{Figures/Figure_1}
%\caption{Placeholder image of Iris with a long example caption to show justification setting.}
%\label{fig:computerNo}
%\end{figure}

@@

\section{Funding}
%TODO the permed ack wasn't right, it should be:
he research leading to these results has received funding from the European Union's Horizon 2020 Programme under the PerMedCoE Project (\url{http://www.permedcoe.eu}), grant agreement n\textsuperscript{o}~951773.
% This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773]
The project was partially supported by Charles University, SVV project number~260698.

>>>>

To showcase the utility of the new implementations, we performed benchmarking on both existing models and large-scale synthetic models. As the main results, MaBoSS.GPU may provided over 200\texttimes\ speedup over the current version of MaBoSS on a wide range of models using contemporary GPU accelerators, and MaBoSS.MPI is capable of almost linear performance scaling with added HPC resources, allowing similar speedups by utilizing the current HPC infrastructures.

@@

\subsection{Computational complexity of parallel MaBoSS algorithm}

@@

We estimate the time required to simulate $c$ trajectories as follows:
For simplification, we assume that a typical Boolean logic formula in a model of $n$ nodes can be evaluated in $\mathcal{O}(n)$ (this is a very optimistic but empirically valid estimate). With that, the computation of all possible transition rates (Algorithm~\ref{alg:iter}, line $2$) can be finished in $\mathcal{O}(n^2)$. The selection of the flipping bit (Algorithm~\ref{alg:iter}, line $4$) can be finished in $\mathcal{O}(n)$, and all other parts of the iteration can finish in $\mathcal{O}(1)$. In total, the time complexity of one iteration is $\mathcal{O}(n^2)$. If we simulate $c$ trajectories with an upper bound of trajectory length $u$, the simulation time is in $\mathcal{O}(c \cdot u \cdot n^2)$.

In an idealized PRAM model with infinite parallelism, we can optimize the algorithm in the following ways:
\begin{itemize}
    \item Given $c$ processors, all trajectory simulations can be performed in parallel, reducing the time complexity to $\mathcal{O}(m \cdot n^2)$. (Note that this does not include the results aggregation. See \emph{Statistics aggregation} section for further description.)
    \item With $n$ processors, the computation of transition rates in the simulation can be done $\mathcal{O}(n)$ time, and the selection of the flipping bit can be done in $\mathcal{O}(\log{n})$ time using a parallel prefix sum, giving $\mathcal{O}(n)$ time for a single iteration.
\end{itemize}
Thus, using a perfect parallel machine with $c \cdot n$ processors, the computation time can be reduced to $\mathcal{O}(u \cdot n)$. Notably, the $\mathcal{O}(u)$ simulation steps that must be performed serially remain a major factor in the whole computation time.

@@

In the CPU version of MaBoSS, the simulation part is the most computationally demanding part, with up to 80\% of MaBoSS runtime spent by just evaluating the Boolean formulas (the exact number depends on the model). The original formula evaluation algorithm in MaBoSS used a recursive traversal of the expression tree, which (apart from other issue) causes memory usage patterns unsuitable for GPUs: the memory required per each core is not achievable in current GPUs, and there are typically too many cache misses~\cite{karlsson2000prefetching}.

@@

For optimizing the statistics aggregation, MaBoSS.GPU heavily relies on the fact that the typical number of non-internal nodes in a real-world MaBoSS model rarely exceeds 10 nodes, regardless of the size of the model. This relatively low number of states generated by non-internal nodes allows us to materialize the whole statistics structure (called ``histogram'') as a fixed-size array (rarely exceeding $2^{10}$ elements).

This approach allows us to avoid storing the states as the keys, and gives a simple approach that can map the state to the histogram index using simple bit masking and shifting instructions. Further, we use several well-known GPU histogram update optimizations to improve the performance, including shared memory privatization and atomic operations.

@@

For the benchmarks, we used $3$ real-world models of $10$, $87$ and $133$ nodes (\textsc{cellcycle}~\cite{faure2006cellcycle}, \textsc{sizek}~\cite{sizek2019boolean} and \textsc{Montagud}~\cite{montagud2022prostate}). In order to test the scalability of the GPU and MPI implementation, we also created several synthetic models with up to $1000$ nodes. Synthetic models were designed in a way such that the length of each trajectory is predictable, and the models have no stable states. Also, the number of non-internal nodes was kept low (10 nodes) to enable the usage of the histogram optimization. The synthetic models together with their python generator are available in the replication package.

The GPU implementation benchmarks were run on a datacenter-grade NVIDIA Tesla A100 GPU and a consumer-grade NVIDIA RTX 3070 Laptop GPU. The CPU implementation benchmarks were run on a 32-core Intel Xeon Gold 6130 CPU with multithreading. The CPU implementation was compiled with GCC 13.2.0, and the GPU implementation was compiled with CUDA 12.2. Each measurement was repeated $10$ times, and the average runtime was used as the final results.

The MPI implementation benchmarks were run on the MareNostrum 4 supercomputer~\footnote{\url{https://www.bsc.es/marenostrum/marenostrum}}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{plots/real.pdf}
    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on real-world models. Each model is simulated with 1 million trajectories.}
    \label{fig:real}
\end{figure}

In Figure~\ref{fig:real}, we compare the wall time of the CPU and GPU implementations on real-world datasets. The GPU implementation is faster than the CPU implementation on all models, and the speedup shows to be more significant on the models with more nodes and longer trajectories. On the \textsc{Montagud} model with 133 nodes, but a relatively short average trajectory, we achieve $145\times$ speedup. On a slightly smaller \textsc{sizek} model with a longer average trajectory, the speedup is up to $326\times$.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{plots/nodes.pdf}
    \caption{Wall time comparison of MaBoSS and MaBoSS.GPU on synthetic models. Each model is simulated with 1 million trajectories. The two panels differ by inclusion of the runtime GPU code compilation, showing its impact on total run time.}
    \label{fig:synth}
\end{figure}

Figure~\ref{fig:synth} shows much finer performance progression on synthetic models. We observed that the CPU variant starts to progress steeper at around the $100$ nodes boundary. We assume that the implementation hits the cache size limit, and the overhead of fetching the required data from the memory becomes dominant. The same can be observed in the GPU variant later at around $200$ nodes. Expectably, the cache-spilling performance penalty is much more significant on GPUs. Overall, the results suggest that the optimization of dividing transition rate computations among multiple threads, as mentioned in \emph{Implementation} section, may provide a better speedup for bigger models, as it alleviates the register and cache pressure.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{plots/nodes-compilation-big-NVIDIA Tesla A100 (MaBoSS.GPU).pdf}
    \caption{The time spent in the runtime compilation of the Boolean formulas simulating models with varying numbers of nodes, trajectories, and formula lengths.}
    \label{fig:comp}
\end{figure}

Additionally, Figure~\ref{fig:synth} shows the total runtime of the GPU implementation including the runtime compilation step. Comparing the panels, we observe that that the relative runtime compilation overhead quickly disappears with increasing model size. Figure~\ref{fig:comp} shows the results of a more detailed benchmarks for this scenario, as run on the NVIDIA Tesla A100 GPU. We observed that the compilation time is linearly dependent on the number of nodes and formula lengths (measured in the number of occurring nodes). Notably, as soon as the simulation becomes more complex (e.g., by increasing the number of nodes or simulated trajectories), the compilation time becomes relatively negligible even for models with unrealistically long formulae. This suggests that the runtime compilation is a viable optimization methodology also for much larger models.

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=3.25in]{plots/sizek_mpi.pdf}
\caption{Scalability results of MPI implementation on Sizek model with 20 cores per MPI node.}
\label{fig:sizek_results}
\end{figure}

Figure~\ref{fig:sizek_results} shows the efficiency of the MaBoSS.MPI implementation on the \textsc{sizek} model. We ran multiple suites, ranging from a single MPI node up to 192 nodes, each running 20 cores. We can observe a close-to-linear speedup of up to 64 MPI nodes (1280 cores), and a plateau for larger suites (Figure~\ref{fig:sizek_results}, green). This can be explained by hitting an expectable bottleneck in parallelization overhead and MPI communication cost when the problem is divided into too many small parts.

\begin{figure}%[tbhp]
\centering
\includegraphics[width=3.25in]{plots/synth_mpi_speedup.pdf}
\caption{Speedup scaling of MPI implementation on a synthetic model with 1000 nodes and 32 cores per MPI node.}
\label{fig:synthetic_results}
\end{figure}

@@

% \begin{figure}%[tbhp]
% \centering
% \includegraphics[width=.8\linewidth]{plots/large_model.pdf}
% \caption{Scalability results for synthetic model with 1000 nodes}
% \label{fig:synthetic_results}
% \end{figure}

%\begin{figure}%[tbhp]
%\centering
%\includegraphics[width=.8\linewidth]{plots/Figure_1}
%\caption{Placeholder image of Iris with a long example caption to show justification setting.}
%\label{fig:computerNo}
%\end{figure}

@@

\section{Funding}
The research leading to these results has received funding from the European Union's Horizon 2020 Programme under the PerMedCoE Project (\url{http://www.permedcoe.eu}), grant agreement n\textsuperscript{o}~951773.
% This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773]
The project was partially supported by Charles University, SVV project number~260698.
